
# Break-Even Analysis {#sec-beq-chapter}

```{r}
#| output: false
#| include: false
library(tidyverse)
#library(readxl)
library(ggthemes)
library(scales)
library(gt)
knitr::opts_chunk$set(echo = TRUE) 
```

## Introduction

A critical task for you as an entrepreneur is to ensure your selling price per unit surpasses your cost per unit. This sounds simple, but it's nuanced. If your venture has no fixed costs, just ensure your price is above your average variable cost. However, most businesses do face fixed costs, meaning your total revenue over a period must exceed all your costs, including these fixed expenses. Remember, covering fixed costs can take time, and your venture needs sufficient revenue growth to meet these expenses.

Break-even analysis helps you determine the break-even quantity - the volume of sales needed to cover all costs without making a profit. Before you reach break-even, you are operating at a loss that you must cover with debt or investor money. After you reach break-even, you can begin paying back the losses you incurred before reaching break-even. Without large amounts of outside funding, it may be necessary to invest in a different cost structure that is more favorable in the short-run even if it leads to higher costs in the long run. If this condition seems unreachable, it's a red flag for your business model.

## Calculating Break-Even Quantity

The key to break-even analysis is to determine how many units you must sell to cover your fixed costs in additions to your variable costs. To find this quantity, set your profit function (revenue minus costs) to zero, assuming you know your profit-maximizing price $\mathsf{P^*}$ . The break-even quantity, denoted as $\mathsf{\underline{Q}}$, is then calculated by dividing your fixed costs by the margin (price minus variable cost). The equation is:

\begin{align}
          \mathsf{\pi} &= \mathsf{P\ Q - f - cQ} = 0 \notag \\[8pt]
\mathsf{\underline{Q}} &= \mathsf{\dfrac{f}{P^*-c}}
\end{align}

which is, in words, the fixed cost divided by the margin.

## Applying Break-Even Analysis: A Practical Example

Imagine you estimated a demand curve from a sample of potential customers:

$$ \mathsf{q = 12 - 0.02 \ p.} $$

Scaling this demand curve up to your target market size is essential. Let's say your potential market is 250,000 customers, scaled up from the 50 in your sample. Your adjusted demand curve becomes:

\begin{aligned}
\mathsf{Q} &= \mathsf{\frac{250,\!000}{50} \ q} \\[8pt]
           &= \mathsf{60,\!000 - 100 P} 
\end{aligned}

Now, using a specific cost structure:

$$ \mathsf{C = 50,\!000 - 200Q} $$

This combination of demand and cost curves leads to a profit-maximizing price of $\mathsf{P^* = \$400}$ and maximized profit at that price is $\mathsf{\pi = \$3,\!950,\!000}$. With the demand curve, cost curve, and profit maximizing price, break-even quantity is a simple calculation.

\begin{align}
\mathsf{\underline{Q}} &= \mathsf{\frac{f}{P^*\! - c} = \frac{50,\!000}{400 - 200} } \\[8pt]
                       &= \mathsf{\frac{50,\!000}{200} }\\[8pt]
                       &= \mathsf{250}.
\end{align}

Before this startup is able to reliably sell 250 units, it is operating at a loss. Every unit of quantity above 250 generates positive profit.

The relationship between revenue, cost, and break-even quantity is illustrated in @fig-beq.

```{r}
#| label: fig-beq
#| echo: false
#| fig-cap: "Break-even quantity where revenue equals cost and positive profits begin"
#| fig-cap-location: margin

###  Break even quantity
f <- 50000
c <- 200

fTC <- function(Q) f + c*Q
fF <- function(Q) f 

Pstar <- 400
fR <- function(Q) Pstar * Q 

rev_bounds <- tibble(x = c(0, 500))    # Create tibble containing plot range for x-axis (P)
eqn1 <- " R = P Q = 400 Q"
eqn2 <- " C = 50,000 + 200 Q "
eqn3 <- " underline(Q) == 250 "
eqn4 <- " f = 50,000 "
ggplot(rev_bounds, aes(x)) +                     # Draw ggplot2 plot
  #
  
  #
  geom_function(fun = fF, color = "#FA7240", linewidth = 2) +
  geom_function(fun = fTC, color = "#DA0406", linewidth = 2) +
  geom_function(fun = fR, color = "RoyalBlue", linewidth = 2) +
  labs(#title = "Revenue for a ticket to a dollar theater",
    x = "Quantity",
    y = "Revenue and Cost") +
  #
  geom_text(x=360, y=170000, label="Revenue", hjust = 1, size = 4, parse=F) +
  geom_text(x=360, y=158000, label=eqn1, hjust = 1, size = 3, parse=F) +
  #
  geom_text(x=450, y=120000, label="Total Cost", hjust = 0.5, size = 4, parse=F) +
  geom_text(x=450, y=108000, label=eqn2, hjust = 0.5, size = 3, parse=F) +
  #
  geom_text(x=260, y=18000, label="Break-even Quantity", hjust = 0, size = 4, parse=F) +
  geom_text(x=260, y=6000, label=eqn3, hjust = 0, size = 3, parse=TRUE) +
  #
  geom_text(x=400, y=68000, label="Fixed Cost", hjust = 0.5, size = 4, parse=F) +
  geom_text(x=400, y=56000, label=eqn4, hjust = 0.5, size = 3, parse=F) +
  #
  #scale_x_continuous(labels=scales::dollar_format()) + 
  scale_y_continuous(lim = c(-0,200000), labels=scales::dollar_format()) + 
  theme_hc() + # base size = 12, font family = sans, ... 
  #  scale_color_brewer(palette="GnBu", direction=-1) +
  theme(legend.justification=c(0, 1), legend.position=c(1.8, 1)) # hides legend outside the frame
```

### Break-even Quantity and Time Determine Your Funding Needs

The relationship between revenue, cost, and break-even quantity, illustrated in @fig-beq tells you the scale needed for profitability. However, it doesn't indicate the time required to reach this scale, which depends on sales growth. Depending on demand, you may expect to sell 250 units every hour or it may take more than a year to sell that many units.

You should estimate how long you think it will take for your sales to reach the break-even quantity based on your demand curve and your early growth rate. That estimate will guide you in how much money you will need from outside sources for your venture to reach profitability. Estimate how much time it'll take for your sales to reach the break-even quantity. This will guide your funding needs to sustain the venture until it becomes profitable.

### Understanding the Impact of Cost Structure

Break-even analysis isn't just a mathematical exercise; it's a strategic tool. It guides your decisions on cost structures and underscores the importance of understanding your market and sales potential. Different cost structures lead to different break-even points and, by extension, different funding requirements. This is detailed in @tbl-beq-cost-structures. Observe how changes in fixed and variable costs influence the break-even quantity, shaping your strategic decisions.

```{r}
#| label: tbl-beq-cost-structures
#| tbl-cap: "Impact of changing cost structure on break-even quantity"
#| tbl-cap-location: margin

# Creating the data frame
cost_structures <- data.frame(
  Cost_Structure = c(1, 2, 3),
  Fixed_Cost = c("$5000", "$50,000", "$500,000"),
  Variable_Cost = c("$300", "$200", "$100"),
  Profit_Maximizing_Price = c("$450", "$400", "$350"),
  Break_Even_Quantity = c(33.33, 250, 2000)  # Retaining as numeric
)

# Creating and styling the gt table
gt(cost_structures) %>%
  cols_label(
    Cost_Structure = "Cost Structure",
    Fixed_Cost = "Fixed Cost",
    Variable_Cost = "Variable Cost",
    Profit_Maximizing_Price = "Profit-maximizing Price",
    Break_Even_Quantity = "Break-even Quantity"
  ) %>%
  fmt_number(
    columns = Break_Even_Quantity,
    decimals = 2,
    drop_trailing_zeros = TRUE
  ) |>
  cols_align_decimal() %>% 
  tab_options(
    column_labels.font.size = "large"
  )

```

### Case Study: Launching a No-Frills Car Brand

<br> **The Modern Automobile: A Technology Showcase**

In today's world, cars are not just about transportation; they're marvels of technology. Imagine the latest models with their array of features: airbags, anti-lock brakes, anti-theft systems, and a suite of automated systems like GPS, on-board WIFI, and self-driving capabilities. Initially, these features are exclusive to luxury cars due to their high development costs. But as technology evolves, even entry-level cars start boasting these sophisticated systems. By 2016, electronics constitute a third of a car's cost, a significant price bump passed on to consumers. For tech enthusiasts, this is a bargain; for those seeking basic transportation, it's a steep price to pay.

**The Market Gap: Simplicity in Complexity**

Amidst this tech escalation, there's a gap: what about consumers who want simple, affordable cars? This overlooked segment could represent a unique entrepreneurial opportunity. What if you could start a company focused on building basic, low-tech cars? Let's dive into the financial feasibility of this idea.

**The Cost of Car-Making**

Developing a new car platform is a costly affair, typically ranging from \$700 million to \$2 billion. Our approach, eschewing expensive tech, might reduce these costs. Let's estimate a development cost of \$500 million, allocated to design, engineering, and factory retooling. The big automakers average a \$2500 profit per car, but for smaller, simpler cars, this margin might shrink to \$2000. To break even, you'd need to sell a quarter of a million cars - a daunting target.

**Reality Check: The Sales Challenge**

In 2015, the U.S. car market saw record-breaking sales of 1.7 million cars. But how realistic is our target of 250,000 cars? Comparing with smaller, established brands like Buick, Acura, or Land Rover offers perspective. These brands sold hundreds of thousands of cars annually, but a new brand like ours would likely start much smaller. Tesla, a relatively new entrant, sold only 22,800 cars in 2015. At that rate, it would take over a decade to recover the development costs.

**The High Stakes of High Fixed Costs**

The auto industry's high fixed-cost structure presents a formidable barrier to new entrants, especially at the low-end market. Tesla's success comes from catering to an affluent customer base, offering high-tech cars at premium prices. Our proposed venture sits at the opposite end: targeting cost-conscious consumers with simpler, cheaper vehicles. The challenge is immense: attract enough buyers willing to trade tech for savings and achieve the massive sales volume needed to cover the fixed costs.

**The Verdict: A Tough Road Ahead**

In conclusion, breaking into the auto industry at the low end poses significant risks. The sheer volume of sales needed to cover fixed costs is daunting, and the competitive landscape is unforgiving. Any entrepreneur eyeing this opportunity must carefully weigh these factors against the potential of tapping into an underserved market.

<br>

## Workout Problems

### Dorsal Packs

A group of entrepreneurs, concerned about pick-pocketing in tourist destinations, created the Dorsal travel pack. This stylish, modern backpack has zippers facing inward, toward the back of the wearer, for security. For added security, it is made of slash-resistant fabric in nondescript colors to avoid the attention that more colorful, fashion-oriented brands draw.

The team validated the problem and solution, refined the concept, and confirmed direct sales through their website. They conducted a profitability analysis with 107 potential customers, who rated the innovation, suggested improvements, stated their maximum willingness to pay, and provided travel and demographic information.

### Dorsal Pack data

Initially, the team reached out to members of their social networks and invited those who had recently traveled or were planning to travel in the near future to help. They were able to get 112 responses from potential customers who had traveled or were about to travel.

#### Demand and Revenue

The respondents were shown the backpack and its features. They were asked to rate the innovation and offer suggestions to improve it. They were also asked "What is the most you would be willing to pay for the Dorsal travel pack?" Finally, they provided information about their travel patterns as well as demographics. Because the travel pack is a durable good for most travelers, they focused on the of willingness to pay from respondents who had traveled or were planning to travel soon. We see the following responses:

```{r}
#| label: dp-data
#| echo: true

dp <- tibble(wtp = c(64, 18, 46, 92, 110, 138, 113, 89, 0, 258, 205, 0, 18, 202, 46, 258, 0, 141, 0, 46, 61, 101, 64, 215, 95, 43, 46, 46, 132, 21, 18, 113, 9, 18, 21, 18, 104, 6, 0, 101, 6, 224, 322, 18, 316, 156, 104, 322, 285, 208, 316, 0, 288, 95, 6, 52, 46, 0, 18, 64, 98, 248, 18, 110, 0, 67, 0, 18, 0, 89, 132, 101, 18, 215, 18, 0, 0, 104, 285, 3, 46, 141, 322, 291, 89, 0, 101, 113, 67, 58, 132, 215, 224, 291, 9, 291, 267, 6, 6, 61, 178, 285, 64, 126, 0, 101, 15))

# inspect the data
glimpse(dp)
```

#### Cost Structure

Dorsal plans to have the pack manufactured by a contract manufacturer who charges \$40 per backpack to be manufactured, packaged, and delivered to the startup's offices. The startup is also spending \$200 per month for web hosting and \$5500 per month for contracted web development. Their total cost curve is linear: $\mathsf{C = cQ + f}$ where $\mathsf{c}$ is the variable cost per unit and $\mathsf{f}$ is fixed cost per month.

#### Rescaling

To conduct profitability analytics, it's crucial to align the scale of sample demand with the fixed costs required to serve the market. For this analysis, we will rescale the demand up to the market level. Concerning market size:

-   **International Travelers from the U.S.:** According to the U.S. National Travel and Tourism Office, approximately 93 million international trips were made by U.S. citizens in 2019. We focus on 2019 data because international travel dynamics significantly changed due to the COVID-19 pandemic, and recent recovery data is not yet fully available.

    -   We estimate that the average international traveler makes about 1.5 trips per year, accounting for both frequent travelers and those who rarely travel abroad.
    -   It is generally estimated that 70 to 80 percent of these trips are for leisure purposes.
    -   Combining these factors, we assume there were approximately 43.4 million unique international travelers in 2019.

-   **Percentage of Young Travelers Likely to Use a Backpack:** While detailed data is unavailable, a conservative estimate suggests that 25% of these travelers are in the 20-39 age group, a demographic typically more active in international travel. This estimation brings our target population down to 10.85 million.

-   **Interest in Security:** Estimating the proportion interested in security-enhanced travel gear is challenging. However, considering increased travel safety awareness, we assume that around 20% of the travelers in this age group might be interested in a more secure backpack. This further narrows our target population to approximately 2.17 million.

-   **Market share in a crowded backpack market:** Recognizing the diverse needs and preferences of travelers, as well as the crowded nature of the backpack market, achieving a 1-5% market share would be substantial. Opting for a midpoint estimate of 3%, we approximate the target market to be around 65,100 potential customers.

In summary, our assumption for the target market is about 65,100 potential travelers, and the Dorsal team has sampled data from 112 individuals within this segment for their analysis.

### Questions

1.  What is the rescaling factor $\mathsf{m}$ based on the calculation of the target population?

2.  How many backpacks would you expect to sell to the sample of 107 respondents at a price of \$110?

3.  How many backpacks would you expect to sell to the complete market at a price of \$110?

4.  What is the fixed cost?

5.  What is the unit variable cost?

6.  What is the *profit maximizing price* for a Dorsal pack? Use the rescaled market demand curve for your calculation and continue to use the market demand for the rest of the questions.

7.  What is the *quantity* of travel packs sold at the profit maximizing price?

8.  What is the *revenue* earned by selling Dorsal packs at the profit maximizing price?

9.  What is the *profit* earned by selling Dorsal packs at the profit maximizing price?

10. What is the *difference* between the profit maximizing price and the revenue maximizing price for Dorsal packs? Specifically calculate this as $\mathsf{P^*_{\pi_{max}} - P^*_{R_{max}}}$.

11. What is the *difference* between the revenue when charging the profit maximizing price and the revenue when charging the revenue maximizing price for Dorsal packs? Specifically, calculate this as $\mathsf{R^*_{\pi_{max}} - R^*_{R_{max}}}$.

12. What is the *difference* between the profit when charging the profit maximizing price and the profit when charging the revenue maximizing price for Dorsal packs. Specifically, calculate this as $\mathsf{\pi^*_{\pi_{max}} - \pi^*_{R_{max}}}$.

13. Should the Dorsal pack team charge the profit maximizing price or the revenue maximizing price? Choose the best price and un-comment it.

    1.  best_price \<- "They should charge the profit maximizing price."
    2.  best_price \<- "They should charge the revenue maximizing price."

14. The Dorsal pack team is seeking funding from investors and wants to be able to communicate how long it will need investor funding to reach profitability. What is the breakeven quantity for the Dorsal pack?\
      *Hint: You can calculate this two different ways:*

    1.  *The simple formula in the reading;*
    2.  *Create a profit function as a function of quantity and find its root.*

<br>
