---
subtitle: "The procedure you run yourself"
---

# The Code Layer: Without a Rival {.unnumbered #sec-code-layer}

This appendix is for the reader who would rather write the code than drive a form or supervise a
model. It performs the same arithmetic as the [Profit Analytics
app](https://rconnect.byu.edu/ProfitAnalytics) and the same procedure as [The Method Layer: Without
a Rival](method-layer.qmd), in R and in Python, with the reasoning kept where a person can read it.

**Why code rather than a prompt.** An earlier version of this material was written as prompts: a
paragraph describing what to build, handed to an AI. It kept collapsing into the same thing. Asked
in words for a transformation that counts every respondent and keeps the zeros, a model would
return something conventional instead — fit a regression, drop the outliers, put the prices on an
even grid. The only prompt that reliably produced the right code was one containing the code, at
which point the prompt was doing nothing. So here is the code. Lift a block, change the variable
names, run it.

**Every block below runs on `muscle-cola.csv`**, the practice dataset used throughout the book: 46
respondents, a protein cola sold through gyms. The printed numbers are what you should see. If
yours differ, the difference is in your setup and it is worth finding before you trust the method
on data whose answer you do not know.

---

## What you need {#sec-cl-data}

Three columns, one row per respondent.

| Column | Meaning | Used by |
|---|---|---|
| `wtp` | the most this respondent would ever pay for one unit | both elicitations |
| `quantity` | how many they would take per period **at that maximum price** | how-many only |
| `quantity_at_P0` | how many they would take per period **if it were free** | how-many only |

**Every screened-in respondent stays in.** Someone whose maximum is zero is not a broken row; they
are a member of your market telling you they will not buy, which is one of the more useful things
anyone will tell you. Someone who would take none even when it is free is the same. The most common
way to get a wrong demand curve is to quietly drop them, and nothing in the output will tell you
that you did.

**The line that matters is not buyer versus non-buyer. It is inside the target population versus
outside it.** A non-buyer inside the population is evidence: the market contains people who decline
at every price, and a curve that hides them overstates demand. A respondent outside the population
is not evidence at all, whichever way they answered — an enthusiastic yes from someone who would
never face this purchase corrupts the estimate exactly as much as their no would. That distinction
is the screener's job, not the analysis's, and it is why a zero from someone who should not have
passed the screen is a sampling problem rather than a demand observation.

**A blank is not a zero.** A typed zero is an answer. A blank is an unanswered question, and
imputing zero for it manufactures a non-buyer who never existed — pushing the curve down and making
the offering look worse than the evidence says.

The exception is a blank your instrument created. **If the form branched** — a respondent said they
would pay nothing and was routed past the quantity questions — the blank *is* a zero, because the
answer is implied by the path they took. They were not asked; they were answered for, correctly.

So the rule is: **a blank means whatever your form made it mean.**

| Where the blank came from | Read it as | Then |
|---|---|---|
| The form branched past the question | zero | keep the respondent |
| The question was optional and they skipped it | missing | drop the respondent from that analysis, and report how many |
| You do not know which | missing | drop, report, and fix the instrument before the next fielding |

**And the design rule that makes this nearly moot:** make the valuation questions required, with
zero explicitly permitted. Then blanks should not occur at all, and any that do came from a branch.

`muscle-cola.csv` has no blanks in these three columns. All 11 respondents who would pay nothing
typed a zero, so nothing in this appendix depends on the rule above — but your own data will.

::: {.panel-tabset}

### R

```{r}
#| label: cl-load-r
#| eval: false
library(dplyr); library(readr); library(tidyr); library(purrr)

d <- read_csv("muscle-cola.csv", show_col_types = FALSE) |>
  select(wtp, quantity, q0 = quantity_at_P0) |>
  mutate(across(everything(), ~ replace_na(as.numeric(.x), 0)))

nrow(d)                 # 46
sum(d$wtp <= 0)         # 11  would pay nothing
sum(d$q0  <= 0)         # 12  would take none even at a price of zero
max(d$wtp)              # 3.51
median(d$wtp[d$wtp > 0])# 1.99
```

### Python

```{python}
#| label: cl-load-py
#| eval: false
import csv, statistics

def num(x):
    x = (x or "").strip()
    return float(x) if x else 0.0          # blank means zero, not missing

rows = list(csv.DictReader(open("muscle-cola.csv")))
wtp  = [num(r["wtp"])             for r in rows]
q    = [num(r["quantity"])        for r in rows]
q0   = [num(r["quantity_at_P0"])  for r in rows]

len(rows)                                  # 46
sum(1 for w in wtp if w <= 0)              # 11
sum(1 for x in q0  if x <= 0)              # 12
max(wtp)                                   # 3.51
statistics.median([w for w in wtp if w > 0])   # 1.99
```

:::

---

## Yes/no demand: counting {#sec-cl-yesno}

When a buyer takes one unit per period or none, the transformation is a count. **At each price, how
many respondents said they would pay at least that much?** No regression, no interpolation, no
assumption about anyone's behaviour beyond the answer they gave.

Two details decide whether your curve is right.

**Use the prices your respondents named.** The distinct willingness-to-pay values are where the
evidence is. An evenly spaced grid you invent puts as much weight on the sparse tail as on the
dense middle, and it changes the answer.

**Add a price of zero to that set.** Respondents whose maximum is zero take one when it is free
and nothing above it, which is exactly what `w >= p` produces. They belong at that point and
nowhere else.

::: {.panel-tabset}

### R

```{r}
#| label: cl-yesno-r
#| eval: false
demand_yes_no <- function(wtp) {
  prices <- sort(unique(c(0, wtp)))          # the respondents' own prices, plus zero
  tibble(price    = prices,
         quantity = map_dbl(prices, \(p) sum(wtp >= p)))
}

yn <- demand_yes_no(d$wtp)
yn |> filter(price %in% c(0, 1, 2))
#> price quantity
#>     0       46
#>     1       32
#>     2       15
```

### Python

```{python}
#| label: cl-yesno-py
#| eval: false
def demand_yes_no(wtp):
    """One unit per buyer. At price p, everyone whose maximum is >= p buys."""
    prices = sorted(set(wtp) | {0.0})
    return [(p, sum(1 for w in wtp if w >= p)) for p in prices]

yn = dict(demand_yes_no(wtp))
yn[0.0], yn[1.0], yn[2.0]                  # 46, 32, 15
```

:::

That table is the demand curve: a staircase that steps down as the price rises. It is not an
estimate yet and nothing has been fitted. Everything in the next section is the same operation with
quantities that can exceed one.

---

## How-many demand: horizontal summation {#sec-cl-howmany}

When a buyer takes several per period, each respondent's three anchors describe their own small
demand line:

- `q0` units at a price of zero
- `quantity` units at their own maximum, `wtp`
- nothing above it — a price beyond what they will pay does not make them buy fewer, it makes them
  stop

Join the first two points with a straight line and you have that person's demand. **Then add the
quantities at each price.** That sum is the market curve, and the word *horizontal* refers to the
direction you are adding in: quantity runs along the horizontal axis, so you are summing sideways
at a fixed height.

**Respondents whose maximum is zero carry no slope.** Dividing by `wtp` would divide by zero, so
they contribute their free quantity at a price of zero and nothing above it. That is a coherent
answer — *I would take three if you gave them away and will not pay for them* — not bad data.

::: {.panel-tabset}

### R

```{r}
#| label: cl-howmany-r
#| eval: false
respondent_q <- function(p, q_free, max_wtp, q_at_max) {
  if (max_wtp <= 0) return(if (p <= 0) q_free else 0)   # no slope exists
  if (p > max_wtp)  return(0)                           # above their ceiling, they stop
  max(0, q_free + ((q_at_max - q_free) / max_wtp) * p)  # linear between the anchors
}

demand_how_many <- function(d, prices) {
  tibble(price    = prices,
         quantity = map_dbl(prices, \(p)
           sum(pmap_dbl(list(d$q0, d$wtp, d$quantity),
                        \(f, m, a) respondent_q(p, f, m, a)))))
}

prices <- sort(unique(c(0, d$wtp)))      # again: their prices, plus zero
hm <- demand_how_many(d, prices)
demand_how_many(d, c(0, 1, 2))
#> price quantity
#>     0   1402.0
#>     1    997.9
#>     2    419.0
```

### Python

```{python}
#| label: cl-howmany-py
#| eval: false
def respondent_line(q_free, max_wtp, q_at_max):
    """That respondent's own demand: q_free when free, q_at_max at their
       maximum, nothing above it."""
    if max_wtp <= 0:
        return lambda p: (q_free if p <= 0 else 0.0)
    slope = (q_at_max - q_free) / max_wtp
    return lambda p: max(0.0, q_free + slope * p) if p <= max_wtp else 0.0

def demand_how_many(rows, prices):
    """Horizontal summation: add the respondents' quantities at each price."""
    lines = [respondent_line(*r) for r in rows]      # r = (q_free, max_wtp, q_at_max)
    return [(p, sum(f(p) for f in lines)) for p in prices]

rows   = list(zip(q0, wtp, q))
prices = sorted(set(wtp) | {0.0})
hm     = dict(demand_how_many(rows, prices))
[round(hm[p], 1) for p in (0.0, 1.0, 2.0)]           # [1402.0, 997.9, 419.0]
```

:::

**Check these three numbers before going further.** 1402.0, 997.9 and 419.0 are what the practice
data produces. If your figures are close but not equal, the usual causes are blanks read as missing
instead of zero, respondents dropped for having a maximum of zero, or a price grid of your own
invention rather than the respondents' own prices.

---

## The two are one method {#sec-cl-unify}

> **Yes/no demand is how-many demand where nobody buys more than one.**

Set every respondent's `q0` and `quantity` to 1 and run the how-many code: each line becomes flat
at one unit up to their maximum and zero above it, the sum becomes a head count, and the head count
is the staircase from the previous section. One transformation, two special cases, which is why the
app handles both without switching technique.

It also means a reader who has the three anchors can always produce the yes/no curve, and a reader
who has only `wtp` cannot produce the how-many one. **Collect the three anchors whenever the
product is plausibly bought more than once per period**; you can always discard information you do
not need.

---

## Fitting a curve {#sec-cl-fit}

The staircase is evidence. A fitted curve is an assumption about the prices nobody was asked
about, and it is what you need in order to compute profit at a price you are considering. Keep the
two separate in your head: **counting is nearly assumption-free; fitting invents behaviour.**

**Fit at the prices the respondents named**, the same set you built the curve on, plus zero. An
evenly spaced grid of your own invention weights the sparse tail — where one or two respondents sit
— as heavily as the dense middle, and it moves the answer.

**Fit all three forms and report all three.** Not to crown a winner on fit, but to see how much your
conclusion depends on the shape you assumed. When the three agree, your decision is robust; when
they diverge, the divergence is the finding.

::: {.panel-tabset}

### R

```{r}
#| label: cl-fit-r
#| eval: false
lin_model <- lm(quantity ~ price, hm)

# Exponential: regress on log quantity. Drop zero quantities first -- log(0) is
# undefined -- and say how many you dropped.
exp_model <- lm(log(quantity) ~ price, filter(hm, quantity > 0))

# Sigmoid: a real non-linear optimiser, not a grid search.
sig_model <- nls(quantity ~ SSlogis(price, Asym, xmid, scal), data = hm)

pred_lin <- function(p) predict(lin_model, newdata = tibble(price = p))
pred_exp <- function(p) exp(predict(exp_model, newdata = tibble(price = p)))
pred_sig <- function(p) predict(sig_model, newdata = tibble(price = p))
```

`SSlogis` is written for curves that rise, so on demand data `scal` comes back negative
(here −0.628). That is the curve bending the right way, not an error.

### Python

```{python}
#| label: cl-fit-py
#| eval: false
import numpy as np
from scipy.optimize import curve_fit

P = np.array([p for p, _ in hm_points])
Q = np.array([q for _, q in hm_points])

coef = np.polyfit(P, Q, 1)                       # Q = a + bP
pred_lin = lambda p: coef[0] * p + coef[1]

m = Q > 0                                        # log of zero is undefined
k, lnA = np.polyfit(P[m], np.log(Q[m]), 1)       # ln Q = ln A + kP
pred_exp = lambda p: np.exp(lnA) * np.exp(k * p)

f = lambda p, L, P0, s: L / (1 + np.exp((p - P0) / s))
par, _ = curve_fit(f, P, Q, p0=[max(Q), np.median(P), (max(P) - min(P)) / 4])
pred_sig = lambda p: f(p, *par)
```

:::

On the practice data, all three fit the same 22 points:

| | shape | q at \$1 | q at \$2 |
|---|---|---|---|
| linear | slope −447.6, crosses zero at \$3.12 | 949.1 | 501.5 |
| exponential | A = 2209.9, k = −0.913 | 887.1 | 356.1 |
| sigmoid | L = 1520.7, midpoint \$1.43 | 1008.5 | 434.9 |

**At \$1 they differ by 121 units — 12% — and at \$2 by 145, which is 29%.** Nothing downstream is
more consequential than this choice, and no fit statistic makes it for you.

## Goodness of fit, on one scale {#sec-cl-fitquality}

Compute fit for all three by **predicting quantity at each observed price and scoring there**. This
is easy to get wrong and the error is invisible.

The exponential was fitted by regressing on `ln Q`, so the R² that falls out of that regression
describes how well the *logs* line up. It runs far higher than the same model's fit to actual
quantities and it is not comparable to the linear model's R². On this dataset the log-scale figure
is **0.913** and the quantity-scale figure is **0.607** — second of three by one measure, last by a
wide margin on the other. The sigmoid, fitted by non-linear least squares, has no variance
decomposition at all, so an R² reported for it is descriptive only.

::: {.panel-tabset}

### R

```{r}
#| label: cl-quality-r
#| eval: false
fit_quality <- function(predict_fn, tb) {
  resid <- tb$quantity - predict_fn(tb$price)      # always on the quantity scale
  c(r2   = 1 - sum(resid^2) / sum((tb$quantity - mean(tb$quantity))^2),
    rmse = sqrt(mean(resid^2)))
}

fit_quality(pred_lin, hm)   # r2 0.970  rmse  73.0
fit_quality(pred_exp, hm)   # r2 0.607  rmse 262.5   <- not summary(exp_model)$r.squared
fit_quality(pred_sig, hm)   # r2 0.993  rmse  34.5
```

### Python

```{python}
#| label: cl-quality-py
#| eval: false
def fit_quality(P, Q, predict):
    resid  = Q - np.array([predict(p) for p in P])
    ss_res = float((resid ** 2).sum())
    ss_tot = float(((Q - Q.mean()) ** 2).sum())
    return {"r2":   1 - ss_res / ss_tot if ss_tot else float("nan"),
            "rmse": float(np.sqrt((resid ** 2).mean()))}

fit_quality(P, Q, pred_lin)   # r2 0.970  rmse  73.0
fit_quality(P, Q, pred_exp)   # r2 0.607  rmse 262.5
fit_quality(P, Q, pred_sig)   # r2 0.993  rmse  34.5
```

:::

**Report RMSE beside R², in units you recognise.** "Off by 73 units a month on average" is a
sentence you can argue with; 0.970 is not.

**Neither statistic chooses the curve.** Choose on behaviour: does quantity fall as price rises
everywhere you care about, does the curve stay non-negative, does it say something possible at a
price you have intuition about, and what does it assume beyond the range of your evidence? A model
that fits slightly worse while implying something sensible is the better guide to an irreversible
commitment.

## What to expect on yes/no data {#sec-cl-yesno-fit}

The same three fits run on a yes/no staircase, with one difference worth knowing in advance.

A logistic needs a flat shoulder at low prices before it bends, so that its ceiling and its bend
can be told apart. Horizontal summation produces that shoulder. **A yes/no staircase does not** — it
falls fastest immediately — so the sigmoid is poorly identified there.

On this dataset's yes/no curve it converges and looks respectable, R² 0.957, and still gets the
end that matters wrong: **its ceiling lands at 39.6 when 46 people were observed at a price of
zero**, understating buyers at the low end by eight. The exponential does worse still, R² 0.165,
because a head count does not decay exponentially. The linear fit, R² 0.941, is the honest
description of a staircase.

So on yes/no data: fit all three, and expect the plain line to be the one you defend.

## Look at it {#sec-cl-plot}

Plot the points and the fitted curve together before you use either. This is the check that catches
what statistics hide: a curve sitting consistently above or below the evidence, one that fits the
middle and misses both ends, or one dragged by two extreme answers.

::: {.panel-tabset}

### R

```{r}
#| label: cl-plot-r
#| eval: false
library(ggplot2)
grid <- tibble(price = seq(0, max(hm$price), length.out = 200)) |>
  mutate(linear = pred_lin(price), exponential = pred_exp(price), sigmoid = pred_sig(price)) |>
  tidyr::pivot_longer(-price, names_to = "model", values_to = "quantity")

ggplot(hm, aes(price, quantity)) +
  geom_point(size = 2) +
  geom_line(data = grid, aes(colour = model)) +
  labs(x = "Price", y = "Quantity per period",
       title = "Demand: respondents' points and three fitted curves")
```

### Python

```{python}
#| label: cl-plot-py
#| eval: false
import matplotlib.pyplot as plt
grid = np.linspace(0, P.max(), 200)
plt.plot(P, Q, "o", label="respondents")
for name, fn in (("linear", pred_lin), ("exponential", pred_exp), ("sigmoid", pred_sig)):
    plt.plot(grid, [fn(p) for p in grid], label=name)
plt.xlabel("Price"); plt.ylabel("Quantity per period"); plt.legend()
plt.title("Demand: respondents' points and three fitted curves")
```

:::

**Cover the curves and look only at the points. Would you have drawn that line yourself?** If not,
the difference between what you would have drawn and what was fitted is the finding, and it is
usually telling you something about the evidence rather than about the model.

---

*The next section rescales the fitted curve from your sample to a reachable population, and then
assembles profit from demand and cost.*
