---
subtitle: "The procedure you run yourself"
---

# The Code Layer: Without a Rival {.unnumbered #sec-code-layer}

This appendix is for the reader who would rather write the code than drive a form or supervise a
model. It performs the same arithmetic as the [Profit Analytics
app](https://rconnect.byu.edu/ProfitAnalytics) and the same procedure as [The Method Layer: Without
a Rival](method-layer.qmd), in R and in Python, with the reasoning kept where a person can read it.

**Why code rather than a prompt.** An earlier version of this material was written as prompts: a
paragraph describing what to build, handed to an AI. It kept collapsing into the same thing. Asked
in words for a transformation that counts every respondent and keeps the zeros, a model would
return something conventional instead — fit a regression, drop the outliers, put the prices on an
even grid. The only prompt that reliably produced the right code was one containing the code, at
which point the prompt was doing nothing. So here is the code. Lift a block, change the variable
names, run it.

**Every block below runs on `muscle-cola.csv`**, the practice dataset used throughout the book: 46
respondents, a protein cola sold through gyms. The printed numbers are what you should see. If
yours differ, the difference is in your setup and it is worth finding before you trust the method
on data whose answer you do not know.

---

## What you need {#sec-cl-data}

Three columns, one row per respondent.

| Column | Meaning | Used by |
|---|---|---|
| `wtp` | the most this respondent would ever pay for one unit | both elicitations |
| `quantity` | how many they would take per period **at that maximum price** | how-many only |
| `quantity_at_P0` | how many they would take per period **if it were free** | how-many only |

**Every respondent stays in.** Someone whose maximum is zero is not a broken row; they are evidence
that part of your market will not pay. Someone who would take none even when it is free is the
same. The most common way to get a wrong demand curve is to quietly drop them, and nothing in the
output will tell you that you did.

Blanks are zeros, not missing values, and reading them as `NA` will silently shrink your sample.

::: {.panel-tabset}

### R

```{r}
#| label: cl-load-r
#| eval: false
library(dplyr); library(readr); library(tidyr); library(purrr)

d <- read_csv("muscle-cola.csv", show_col_types = FALSE) |>
  select(wtp, quantity, q0 = quantity_at_P0) |>
  mutate(across(everything(), ~ replace_na(as.numeric(.x), 0)))

nrow(d)                 # 46
sum(d$wtp <= 0)         # 11  would pay nothing
sum(d$q0  <= 0)         # 12  would take none even at a price of zero
max(d$wtp)              # 3.51
median(d$wtp[d$wtp > 0])# 1.99
```

### Python

```{python}
#| label: cl-load-py
#| eval: false
import csv, statistics

def num(x):
    x = (x or "").strip()
    return float(x) if x else 0.0          # blank means zero, not missing

rows = list(csv.DictReader(open("muscle-cola.csv")))
wtp  = [num(r["wtp"])             for r in rows]
q    = [num(r["quantity"])        for r in rows]
q0   = [num(r["quantity_at_P0"])  for r in rows]

len(rows)                                  # 46
sum(1 for w in wtp if w <= 0)              # 11
sum(1 for x in q0  if x <= 0)              # 12
max(wtp)                                   # 3.51
statistics.median([w for w in wtp if w > 0])   # 1.99
```

:::

---

## Yes/no demand: counting {#sec-cl-yesno}

When a buyer takes one unit per period or none, the transformation is a count. **At each price, how
many respondents said they would pay at least that much?** No regression, no interpolation, no
assumption about anyone's behaviour beyond the answer they gave.

Two details decide whether your curve is right.

**Use the prices your respondents named.** The distinct willingness-to-pay values are where the
evidence is. An evenly spaced grid you invent puts as much weight on the sparse tail as on the
dense middle, and it changes the answer.

**Add a price of zero to that set.** Respondents whose maximum is zero take one when it is free
and nothing above it, which is exactly what `w >= p` produces. They belong at that point and
nowhere else.

::: {.panel-tabset}

### R

```{r}
#| label: cl-yesno-r
#| eval: false
demand_yes_no <- function(wtp) {
  prices <- sort(unique(c(0, wtp)))          # the respondents' own prices, plus zero
  tibble(price    = prices,
         quantity = map_dbl(prices, \(p) sum(wtp >= p)))
}

yn <- demand_yes_no(d$wtp)
yn |> filter(price %in% c(0, 1, 2))
#> price quantity
#>     0       46
#>     1       32
#>     2       15
```

### Python

```{python}
#| label: cl-yesno-py
#| eval: false
def demand_yes_no(wtp):
    """One unit per buyer. At price p, everyone whose maximum is >= p buys."""
    prices = sorted(set(wtp) | {0.0})
    return [(p, sum(1 for w in wtp if w >= p)) for p in prices]

yn = dict(demand_yes_no(wtp))
yn[0.0], yn[1.0], yn[2.0]                  # 46, 32, 15
```

:::

That table is the demand curve: a staircase that steps down as the price rises. It is not an
estimate yet and nothing has been fitted. Everything in the next section is the same operation with
quantities that can exceed one.

---

## How-many demand: horizontal summation {#sec-cl-howmany}

When a buyer takes several per period, each respondent's three anchors describe their own small
demand line:

- `q0` units at a price of zero
- `quantity` units at their own maximum, `wtp`
- nothing above it — a price beyond what they will pay does not make them buy fewer, it makes them
  stop

Join the first two points with a straight line and you have that person's demand. **Then add the
quantities at each price.** That sum is the market curve, and the word *horizontal* refers to the
direction you are adding in: quantity runs along the horizontal axis, so you are summing sideways
at a fixed height.

**Respondents whose maximum is zero carry no slope.** Dividing by `wtp` would divide by zero, so
they contribute their free quantity at a price of zero and nothing above it. That is a coherent
answer — *I would take three if you gave them away and will not pay for them* — not bad data.

::: {.panel-tabset}

### R

```{r}
#| label: cl-howmany-r
#| eval: false
respondent_q <- function(p, q_free, max_wtp, q_at_max) {
  if (max_wtp <= 0) return(if (p <= 0) q_free else 0)   # no slope exists
  if (p > max_wtp)  return(0)                           # above their ceiling, they stop
  max(0, q_free + ((q_at_max - q_free) / max_wtp) * p)  # linear between the anchors
}

demand_how_many <- function(d, prices) {
  tibble(price    = prices,
         quantity = map_dbl(prices, \(p)
           sum(pmap_dbl(list(d$q0, d$wtp, d$quantity),
                        \(f, m, a) respondent_q(p, f, m, a)))))
}

prices <- sort(unique(c(0, d$wtp)))      # again: their prices, plus zero
hm <- demand_how_many(d, prices)
demand_how_many(d, c(0, 1, 2))
#> price quantity
#>     0   1402.0
#>     1    997.9
#>     2    419.0
```

### Python

```{python}
#| label: cl-howmany-py
#| eval: false
def respondent_line(q_free, max_wtp, q_at_max):
    """That respondent's own demand: q_free when free, q_at_max at their
       maximum, nothing above it."""
    if max_wtp <= 0:
        return lambda p: (q_free if p <= 0 else 0.0)
    slope = (q_at_max - q_free) / max_wtp
    return lambda p: max(0.0, q_free + slope * p) if p <= max_wtp else 0.0

def demand_how_many(rows, prices):
    """Horizontal summation: add the respondents' quantities at each price."""
    lines = [respondent_line(*r) for r in rows]      # r = (q_free, max_wtp, q_at_max)
    return [(p, sum(f(p) for f in lines)) for p in prices]

rows   = list(zip(q0, wtp, q))
prices = sorted(set(wtp) | {0.0})
hm     = dict(demand_how_many(rows, prices))
[round(hm[p], 1) for p in (0.0, 1.0, 2.0)]           # [1402.0, 997.9, 419.0]
```

:::

**Check these three numbers before going further.** 1402.0, 997.9 and 419.0 are what the practice
data produces. If your figures are close but not equal, the usual causes are blanks read as missing
instead of zero, respondents dropped for having a maximum of zero, or a price grid of your own
invention rather than the respondents' own prices.

---

## The two are one method {#sec-cl-unify}

> **Yes/no demand is how-many demand where nobody buys more than one.**

Set every respondent's `q0` and `quantity` to 1 and run the how-many code: each line becomes flat
at one unit up to their maximum and zero above it, the sum becomes a head count, and the head count
is the staircase from the previous section. One transformation, two special cases, which is why the
app handles both without switching technique.

It also means a reader who has the three anchors can always produce the yes/no curve, and a reader
who has only `wtp` cannot produce the how-many one. **Collect the three anchors whenever the
product is plausibly bought more than once per period**; you can always discard information you do
not need.

---

*The next section fits curves to these points: linear, exponential, and sigmoid, with goodness of
fit computed on the original quantity scale.*
