---
title: "Cost"
subtitle: "What you choose, and what you commit to"
---

```{r}
#| output: false
#| include: false
library(tidyverse)
#library(readxl)
library(scales)
library(gt)
knitr::opts_chunk$set(echo = TRUE) 
source("R/theme-itwd.R")
```
Now that demand is on firmer ground, the analysis changes character.

Until this point the central unknown has been outside you. How customers respond to price had to be learned from the world, because no amount of deciding would produce it. Cost is different. Costs are not revealed by the market; they follow from how you design the thing, who you buy from, what you build rather than rent, and how fast you commit.

That does not make them trivial. It makes them **decisional**, which is a different kind of problem and calls for a different kind of thinking. The goal here is not to forecast expenses accurately. It is to see how a cost structure meets a demand curve, and what that meeting implies about whether the thing is worth doing.

## Why Cost Comes After Demand

Entrepreneurs usually want to reason about cost first, and the instinct is understandable. Costs feel concrete and controllable in a way demand does not.

The trouble is that cost analysis floats when it comes first. You estimate a per-unit cost without knowing how many units. You argue about a lease without knowing whether any price can support it. Precision accumulates in a spreadsheet while the uncertainty that actually decides the question sits untouched.

Once demand is a price-quantity relationship, cost questions acquire anchors. Some costs turn out to matter enormously and others hardly at all, and you can finally tell which is which.

Demand tells you whether revenue is possible. Cost tells you whether that revenue can carry the commitments you are contemplating. Neither answers the question alone.

## What One More Unit Costs You

The first cost question worth asking is deceptively plain.

> *What has to happen, operationally, for one more customer to be served?*

Not what should happen in a mature company, and not what might happen eventually. What actually happens now, given how you have designed this. If somebody buys and almost nothing happens, your variable cost is near zero. If a great deal happens, the cost is real even when you cannot yet price it precisely.

::: {.column-margin .def-margin}
**Variable cost** — any cost triggered by serving one more unit of demand, whatever an accountant would call it.
:::

That question is better than the accounting definition because it looks for **triggers rather than labels**. A **variable cost** is any cost set in motion by serving one more unit. Some costs that feel operational turn out not to be triggered at the margin at all; others that feel minor turn out to be triggered every single time, and multiply into something decisive.

This distinction earns its keep in a specific way. Early ventures rarely fail because someone estimated a per-unit cost a few percent wrong. They fail because a cost was never noticed. Walking through what physically happens when an order arrives finds the missed ones; consulting a category list does not.

Variable cost then does one job that nothing else can do. It sets a hard floor under viable pricing. Below it, selling more makes things worse rather than better, and demand at those prices is economically irrelevant however real it is. So before asking which price is best, ask which prices are even worth considering.

Estimate in ranges rather than points. The purpose is discrimination, not accuracy: costs clearly low enough to work at plausible prices, costs clearly too high, and the handful in between where a small error would change what you do. Only the third group deserves more effort.

And treat variable cost as a choice rather than a fact of nature. How the thing is built, what is outsourced, what quality you target, how you deliver: each is a decision, and each produces a different variable cost against the same demand.

## What You Commit To Before Anyone Buys

Fixed costs answer a different question, and the difference is the whole point.

> *What do I have to commit to before knowing how many customers will buy?*

::: {.column-margin .def-margin}
**Fixed cost** — a cost you incur because you decided to proceed, rather than because a customer decided to buy.
:::

If a cost arrives because a customer bought, it is variable. If it arrives because **you** decided to move forward, it is fixed. That test matters more than any accounting category, because it identifies who is carrying the risk when demand disappoints. A **fixed cost** is best read as a commitment rather than an expense.

What counts is not whether a cost is fixed forever but whether it is hard to reverse, slow to unwind, or binding across the decision you are actually making. A lease, salaried people, specialised equipment, a platform contract, money spent on compliance or brand before anyone has bought anything.

Commitments change the shape of the problem by introducing a threshold. With variable costs alone, every sale above cost helps. Once a fixed cost exists, the venture has to generate enough contribution to justify it before anything good happens at all, and that *enough* is a demand requirement you did not previously have.

Low commitments buy flexibility. Demand can be tested, prices adjusted, the whole thing abandoned without ruin. High commitments spend that flexibility for efficiency: demand now has to be large and dependable, pricing errors get expensive, and early mistakes become difficult to undo. This is usually the real source of risk in a pre-revenue decision, rather than uncertainty about what a unit costs.

Which is why **when** matters more than **how much**. Some commitments must be made now and others can wait until demand is better understood; some can be staged. Delaying preserves the option to learn before binding yourself. Experienced entrepreneurs often look conservative about fixed costs while being genuinely optimistic about demand, and the two are consistent: they are not doubting the market, they are refusing to pay for certainty they do not have yet.

Fixed costs are chosen too. How fast to scale, what capability to build, own or rent, automate or do by hand. Different answers, different exposure, same demand.

## The Shape of a Cost Structure

Put the two together and total cost has a simple form:

$$ \mathsf{C(q)} = \mathsf{f + cq} $$

Total cost is what you committed to before anyone bought, plus what each sale triggers. This is not a theory of cost. It separates two kinds of decision that behave differently, and a cost structure is just a particular choice of $\mathsf{f}$ and $\mathsf{c}$.

That choice does more than move profit up or down. It decides how much risk you are absorbing.

Two ventures can face identical demand at identical prices and present completely different decisions. The high-fixed, low-variable structure looks better whenever demand shows up, because margins are wide and profit accumulates fast. The condition attached to that upside is easy to skip past: demand has to be large enough and reliable enough. High fixed costs concentrate risk, making outcomes acutely sensitive to errors in your demand estimate, your price, or your timing. Low fixed costs spread it, capping the upside and also the damage.

@fig-cost-structures shows several ways of serving the same demand, each a different commitment profile.

```{r}
#| label: fig-cost-structures
#| echo: false
#| warning: false
#| fig-cap: "Cost Structure Is a Choice About Risk: Each curve shows total cost C(q) = f + cq for a different combination of fixed and variable costs. No structure is universally better — each becomes preferable only at sufficient scale."
#| fig-cap-location: margin

#% Cost structure choice depends on scale: different fixed/variable mixes dominate at different quantities."

library(tidyverse)
library(scales)

# Cost structures (edit freely)
F1 <- 5000
F2 <- 50000
F3 <- 500000

c1 <- 300
c2 <- 200
c3 <- 100

fTC1 <- function(Q) F1 + c1 * Q
fTC2 <- function(Q) F2 + c2 * Q
fTC3 <- function(Q) F3 + c3 * Q

# Crossing points (where TC_i == TC_j)
Q12 <- (F2 - F1) / (c1 - c2)
Q23 <- (F3 - F2) / (c2 - c3)

# y-values at crossings for vertical guide lengths
TC12 <- fTC1(Q12)
TC23 <- fTC2(Q23)

# Plot range
Qmax <- 7000
rev_bounds <- tibble(x = c(0, Qmax))

# Helper data for direct labels placed near the right edge
labels_df <- tibble(
  Q = c(3600, 3100, 6300),
  label = c("Cost structure 1\n(low f, high c)",
            "Cost structure 2\n(moderate f, moderate c)",
            "Cost structure 3\n(high f, low c)"),
  TC = c(fTC1(3600), fTC2(3100), fTC3(6300))
)

ggplot(rev_bounds, aes(x)) +
  # Vertical guides at crossing points
  geom_segment(aes(x = Q12, y = 0, xend = Q12, yend = TC12),
               linetype = "dotted", linewidth = 0.5, color = "black") +
  geom_segment(aes(x = Q23, y = 0, xend = Q23, yend = TC23),
               linetype = "dotted", linewidth = 0.5, color = "black") +

  # Total cost curves (grayscale + distinct linetypes)
  geom_function(fun = fTC1, linewidth = 1.8, color = "#FB923C", linetype = "solid") +
  geom_function(fun = fTC2, linewidth = 1.8, color = "#C2410C", linetype = "solid") + # longdash
  geom_function(fun = fTC3, linewidth = 1.8, color = "#7F1D1D", linetype = "solid") + # dotdash

  # Direct labels on curves (avoids legend problems)
  geom_label(data = labels_df, aes(x = Q, y = TC, label = label),
             hjust = 0.5, size = 3, label.size = 0.2, fill = "white") +

  # Crossing annotations (kept minimal and readable)
  annotate("label", x = Q12 + 250, y = 40000,
           label = paste0("Crossing at Q ≈ ", comma(round(Q12))),
           hjust = 0, label.size = 0.2, size = 3, fill = "white") +
  annotate("label", x = Q23 + 250, y = 40000,
           label = paste0("Crossing at Q ≈ ", comma(round(Q23))),
           hjust = 0, label.size = 0.2, size = 3, fill = "white") +

  labs(x = "Quantity", y = "Total cost") +
  scale_y_continuous(labels = dollar_format()) +
  coord_cartesian(xlim = c(0, Qmax), ylim = c(0, 1200000), expand = FALSE) +
  theme_itwd()
```
Where those lines cross is not a target. It is a condition. Picking the high-commitment structure is a bet that demand will arrive at that scale and stay there; picking the low-commitment one keeps you solvent if it does not.

## Break-Even Is a Constraint, Not a Target

Break-even gets talked about as a milestone, which quietly misdescribes it.

It answers one narrow question: how much has to sell before this cost structure stops losing money. That is a constraint you must clear, not an achievement. Reaching it does not mean a venture is attractive, or scalable, or worth what it cost to get there. It means the losses stopped growing. A business that barely breaks even has not succeeded; it has survived.

The useful thing break-even tells you is how demanding your commitments are. High fixed costs push the required quantity up and expose you further to being wrong about demand. Low fixed costs bring it down, usually by accepting a higher cost per unit later.

So break-even is not something to aim at. It is something to inspect, because it converts your cost structure into a statement about how much demand must exist before any of this works, which you can then hold against the demand curve you actually estimated.

## What Cost Cannot Decide

Cost analysis feels decisive in a way it is not. Lower costs feel safer, leaner operations feel prudent, and cutting a commitment feels like progress.

But cost creates no value. It generates no demand. It has no opinion about whether anyone will buy. All it tells you is what serving demand would take, if demand appears.

Which means a cheaper structure can still be the wrong one when demand is small or price-sensitive, and an expensive structure can be right when demand is strong enough to carry it. Cost narrows the field of viable decisions. It does not choose among them, and treating a cost reduction as a result is how ventures optimise their way into a business nobody wants.

Demand and cost only mean something together, and what they mean together is profit.

## Putting It to Work

::: {.reflect}
**Ask yourself — what fires, and what binds?**

Two lists, and they answer different questions.

First: walk through what physically happens when one more customer buys. Somebody packs something, a fee is charged, a message is sent, an hour is spent. Write down each thing that fires. That list is your variable cost, and the point of writing it is to catch what you would otherwise never have noticed.

Second: write down everything you would still owe at the end of the month if nobody bought at all. Rent, salaries, subscriptions, the equipment payment. Those are commitments, and they bind whether or not you were right about demand.

Now look at the second list and mark each item with when you actually have to commit to it. Anything you could postpone until you know more about demand is an option you currently hold, and options are worth keeping.
:::

::: {.key-action}
**The move:** Ask what fires when somebody buys, and what binds when nobody does. Those two questions find more real costs than any category list.
:::

Neither list decides anything on its own. Put them against the demand curve and the question stops being about cost at all: whether what customers will pay, at the quantities they will buy, covers what you had to commit before you knew either.
