---
title: "Scale and Population"
subtitle: "How many customers your commitments require you to find"
---

```{r}
#| output: false
#| include: false
library(tidyverse)
#library(readxl)
library(scales)
library(gt)
knitr::opts_chunk$set(echo = TRUE) 
source("R/theme-itwd.R")
```

Once costs are understood as commitments, a constraint appears that demand curves cannot supply on their own.

A demand curve tells you how the people you studied respond to price. It says nothing about how many such people exist, how many you could ever reach, or how many would have to buy before your commitments are covered. While you had made no commitments, none of that mattered. The moment you have fixed costs, it decides everything.

This chapter is about the arithmetic those commitments impose. Not market sizing as a marketing exercise, but scale as a condition: given what you have agreed to pay whether or not anyone buys, how many customers must you actually find?

## Market Size Is Not Demand

Entrepreneurs talk about market size as though it were part of demand. It is a different quantity answering a different question, and confusing the two hides the one that matters.

Demand describes how customers respond to price: whether quantity is sensitive or stubborn, whether cutting the price brings people in. Scale asks how many people could show up at all. A venture can face intense demand from a small population, or listless demand from an enormous one, and the shape of the curve tells you nothing about which situation you are in.

::: {.column-margin .def-margin}
**Population** — the number of people who could plausibly have the problem. A ceiling on demand rather than a forecast of it.
:::

Treat **population** as a constraint rather than an estimate. It is a ceiling on how much demand could ever materialize, a limit on how much contribution your fixed costs can draw from, and a bound on how wrong an optimistic projection is allowed to be. Population does not tell you what will happen. It tells you what cannot.

This is also why the chapter sits here rather than earlier. With no commitments, population is close to irrelevant; you can serve whoever turns up. With commitments, it becomes decisive, because the commitments have to be covered by however many people exist to cover them.

It is worth saying plainly that this is not market sizing in the usual sense. We are not estimating a share of an imagined market or forecasting growth, which is the move [chapter 2](familiar-tools.qmd) took apart. We start from demand you learned, costs you chose, and commitments you are considering, and then ask whether a large enough reachable population exists to support them. Scale here is a test rather than an aspiration.

## Existing Is Not Being Reachable

A population can exist and still be useless to you.

Existence asks who could plausibly have the problem. Access asks which of those people you can find, talk to, earn trust from, and serve, repeatedly, at a cost you can bear. The second question is answered by structure rather than effort: whether channels exist, whether gatekeepers stand in the way, what regulation applies, how much switching costs bind people to what they already use, whether the social risk of trying something new is high.

*A population that exists but cannot be reached is economically equivalent to one that does not exist.*

Entrepreneurs tend to treat access as a marketing problem to be solved later with more spend. Some access barriers do yield to time and money. Others are structural and will not, and the difference has to be assessed before commitments are justified rather than after.

Your early demand experiments do not settle this. They were run with people who were easy to find, willing to talk, and already aware of the problem, which is exactly right for learning demand and no evidence at all about reach. Recall that in [executing the experiment](executing-experiments.qmd) the response rate itself was the first read on what acquisition costs; this is the same finding arriving with consequences attached.

So access shrinks the population you are allowed to use. Not the number who might benefit, but the number who are reachable, engageable, and realistically serviceable. That is not pessimism. It is the only version of the number that can carry a commitment.

## Adoption Is Not Penetration

One more distinction, and it is the one that trips up otherwise careful people.

Adoption is a customer-level decision: whether this person, facing this price, says yes. It is governed by demand, and your demand curve already contains it. Penetration is an aggregate outcome: what fraction of a population ends up adopting. It is the cumulative result of awareness, access, timing, trust, substitutes, and eventually competition, and none of that appears in a demand curve.

Fixed costs do not care who adopts. They care how many do.

Which is why early evidence cannot give you a penetration rate, however encouraging it looks. *Thirty percent of the people we interviewed said they would buy* is a statement about thirty percent of the people you could find, who were more motivated, better informed, more tolerant of friction, and more forgiving than the population behind them. That does not make the evidence useless. It makes it **local**.

Scaling a local result up to a population is the error in @fig-scaling-mismatch-1, where demand measured on a sample is asked to carry costs incurred at population scale and revenue never catches up. @fig-scaling-mismatch-2 puts the two on the same footing. Nothing about the business changed between those panels. Only the units did, and the first version looks like a verdict.

```{r}
#| label: fig-scaling-mismatch
#| echo: false
#| warning: false
#| layout-ncol: 2
#| cap-location: bottom
#| fig-cap:
#|   - "The scaling mistake: sample demand is asked to carry population-level fixed costs (revenue never catches cost)."
#|   - "Demand on the correct scale: rescaling demand to the target population puts revenue and cost on comparable terms."

library(ggplot2)
library(scales)

# -----------------------------
# Demand (sample) as Q(P)
# -----------------------------
# Linear demand in *quantity as a function of price* (your convention)
a <- 2500
b <- -2
Q_sample <- function(P) pmax(0, a + b * P)   # clamp at 0 so it doesn't go negative

# -----------------------------
# Scale factor: population vs sample
# -----------------------------
# Think of this as (accessible target population / sample size) times any other scaling logic you use.
k <- 12
Q_pop <- function(P) k * Q_sample(P)

# -----------------------------
# Cost structure (population-level fixed cost + per-unit variable cost)
# -----------------------------
F <- 1000000
c <- 200
C_of_Q <- function(Q) F + c * Q

# Express cost as a function of price using demand: C(P) = F + c*Q(P)
C_sample <- function(P) C_of_Q(Q_sample(P))
C_pop    <- function(P) C_of_Q(Q_pop(P))

# -----------------------------
# Revenue as a function of price: R(P) = P * Q(P)
# -----------------------------
R_sample <- function(P) P * Q_sample(P)
R_pop    <- function(P) P * Q_pop(P)

# Plot range for price
P_min <- 0
P_max <- 1300

# Shared formatting helper (keeps visual grammar consistent)
base_theme <- theme_itwd()

# -----------------------------
# Panel A: Scaling mistake
# -----------------------------
plot_scaling_mistake <- ggplot() +
  geom_function(fun = R_sample, xlim = c(P_min, P_max),
                linewidth = 2, color = itwd_demand) +
  geom_function(fun = C_sample, xlim = c(P_min, P_max),
                linewidth = 2, color = "#C2410C") +
  labs(x = "Price", y = "Dollars") +
  scale_x_continuous(labels = label_dollar(), n.breaks = 5, limits = c(0,1249)) +
  scale_y_continuous(labels = label_dollar(), limits = c(0,10000000)) +
  coord_cartesian(xlim = c(P_min, P_max), expand = FALSE) +
  base_theme +
  annotate("label",
           x = 850, y = max(C_sample(seq(P_min, P_max, length.out = 200))) + 1000000 * 0.92,
           label = "Sample revenue\nvs population fixed costs",
           size = 4, label.size = 0.2, fill = "white")

# -----------------------------
# Panel B: Demand scaled to population
# -----------------------------
plot_scaled_correctly <- ggplot() +
  geom_function(fun = R_pop, xlim = c(P_min, P_max),
                linewidth = 2, color = itwd_demand) +
  geom_function(fun = C_pop, xlim = c(P_min, P_max),
                linewidth = 2, color = "#C2410C") +
  labs(x = "Price", y = "Dollars") +
  scale_x_continuous(labels = label_dollar(), n.breaks = 5, limits = c(0,1249)) +
  scale_y_continuous(labels = label_dollar(), limits = c(0,10000000)) +
  coord_cartesian(xlim = c(P_min, P_max), expand = FALSE) +
  base_theme +
  annotate("label",
           x = 850, y = max(C_pop(seq(P_min, P_max, length.out = 200))) * 0.92,
           label = "Revenue and cost\non comparable scale",
           size = 4, label.size = 0.2, fill = "white")

plot_scaling_mistake
plot_scaled_correctly
```
## The Penetration Your Commitments Require

Here is the part that turns all of this into a number.

Penetration usually gets talked about aspirationally. *If we get just five percent of the market.* That framing quietly reverses the logic, because penetration is not something you aim at. It is something your cost structure **demands**, and you can calculate it before you have any idea whether it is achievable.

Every sale contributes what is left after the cost of making it, which is the price minus the variable cost. Divide your fixed commitment by that contribution and you have the number of units required before anything is covered:

$$ \mathsf{q_{required}} = \frac{\mathsf{f}}{\mathsf{p - c}} $$

Divide that by the population you can actually reach, and you have the share of them who must buy.

Work it once. Commit \$50,000. Charge \$40 against a variable cost of \$25, so each sale contributes \$15. You need 3,334 sales. If the population you can genuinely reach is 20,000 people, you need one in six of them, which is an extraordinary number for a new offering nobody has heard of. If it is 200,000, you need one in sixty, which is merely ambitious.

Same commitment, same price, same cost. The only thing that changed was who you can reach, and it moved the requirement by a factor of ten.

```{ojs}
//| echo: false
viewof fCost = Inputs.range([5000, 250000], {value: 50000, step: 5000, label: "Commitment"})
viewof pPrice = Inputs.range([10, 120], {value: 40, step: 1, label: "Price"})
viewof cUnit = Inputs.range([1, 100], {value: 25, step: 1, label: "Cost per unit"})
viewof popn = Inputs.range([2000, 400000], {value: 20000, step: 2000, label: "Reachable population"})
```

```{ojs}
//| echo: false
contrib = pPrice - cUnit
qReq = contrib > 0 ? fCost / contrib : Infinity
penReq = qReq / popn
fmt = d3.format(",.0f")
pct = d3.format(".2%")

html`<div style="border-left:3px solid #146B45;padding:.6rem 0 .6rem .9rem;margin:.6rem 0">
${contrib <= 0
 ? `<p style="margin:0;color:#A33A2A"><strong>No price above cost.</strong> Every sale loses
    money, so no quantity and no population can cover the commitment.</p>`
 : `<p style="margin:0 0 .35rem">Each sale contributes <strong>$${fmt(contrib)}</strong>.
    Covering the commitment takes <strong>${fmt(qReq)}</strong> sales.</p>
    <p style="margin:0">Out of ${fmt(popn)} reachable people, that is
    <strong style="font-size:1.15em">${pct(penReq)}</strong> of them —
    about <strong>one in ${fmt(1/penReq)}</strong>.
    ${penReq > 1
      ? ` <span style="color:#A33A2A">More people than you can reach. This cannot work at
          this price with this commitment.</span>`
      : penReq > 0.20
        ? ` <span style="color:#A33A2A">A fifth of everyone reachable is a very hard
            number for something new.</span>`
        : penReq > 0.05
          ? ` <span style="color:#8A6A12">Demanding, and worth being honest about.</span>`
          : ` <span style="color:#1B6E45">Modest relative to the population.</span>`}</p>`}
</div>`
```

Move the sliders and notice which one has the most leverage. Cutting the commitment lowers the requirement proportionally. Raising the price does too, and faster than most people expect, because it widens contribution rather than merely adding revenue. Enlarging the reachable population lowers the *share* without touching the *count*, which is why the honest version of that number is the one filtered by access rather than the one from the industry report.

Notice also what the calculation is not. It is not a forecast. Nobody is claiming this penetration will occur. It answers a conditional: *if this venture is to work, how much demand must ultimately materialize?* That question comes before forecasting, before competition, and before any conversation about market share.

Which reframes the question you should be asking. Not *is this a big market*, which invites the answer you want, but *is it plausible that this many reachable people will adopt, at a price that clears cost, in time to cover what I have already committed to?* You can hold that against what you know about reach and trust and how new the idea is, and get an answer you would be willing to defend.

Scale, then, is not a strategy. It is not chosen because you want to grow. It is inherited from the commitments you make, and scaling never rescues a structure that does not work; it only amplifies whatever is already there.

## Putting It to Work

::: {.reflect}
**Ask yourself — how many, out of how many?**

Take the commitment you are actually contemplating and run the arithmetic once, by hand. Divide it by the contribution one sale makes, and you have the number of customers you need.

Now the harder half. Write down the population you can genuinely reach: not everyone with the problem, but the ones you could find, contact, earn trust from, and serve again next month at a cost you can carry. If your honest answer is a range, use the low end.

Divide, and say the result out loud as a fraction. One in six. One in forty. One in three hundred. Then ask somebody who knows the market whether a new and unknown offering takes that share of the people it can reach.

If the number embarrasses you, you have learned something before spending anything, which is the cheapest way this lesson is ever available.
:::

::: {.key-action}
**The move:** Calculate the penetration your commitments require before you decide whether it is achievable. The requirement is arithmetic; only its plausibility is a judgment.
:::

You now know how customers respond to price, what serving them costs, what you would be committing to, and how many people would have to buy for that commitment to make sense. What remains is to put those together and see what comes out, which is the question the whole book has been circling.
