---
title: "Moving First, and Meaning It"
subtitle: "Commitment works by removing your own options, which is why it is dangerous"
---

```{r}
#| output: false
#| include: false
library(tidyverse)
library(scales)
library(gt)
knitr::opts_chunk$set(echo = TRUE)
source("R/theme-itwd.R")
```

Both games so far assumed the two firms decided at the same moment, each ignorant of the other. That is a reasonable model of sealed bids and simultaneous launches, and it is a poor model of most competition.

Firms watch each other. One acts, the other sees it and responds. And sometimes a firm acts *specifically* so that the other has to respond to something already done and impossible to undo. That last move is the subject of this chapter, and it is the only reliable way out of the trap the last chapter described.

## Order Changes Outcomes

When one firm moves first, the game is no longer a matrix. It is a sequence, and the right picture is a tree.

```{r}
#| label: fig-two-stage-tree
#| echo: false
#| warning: false
#| fig-cap: "The two-stage structure of every decision in this part. The commitment is chosen first and cannot be unchosen; prices settle afterwards, conditional on it. Working out the payoff at the end of each branch is the only way to choose at the start."
#| fig-align: center
#| fig-width: 8
#| fig-height: 3.6

seg <- tibble::tribble(
  ~x, ~y, ~xend, ~yend,
  0, 0,  1, 0.9,
  0, 0,  1, -0.9,
  1, 0.9, 2, 1.35,
  1, 0.9, 2, 0.45,
  1, -0.9, 2, -0.45,
  1, -0.9, 2, -1.35
)
lab <- tibble::tribble(
  ~x, ~y, ~t, ~h,
  -0.06, 0,    "Hogi Yogi", 1,
  0.5, 0.62,   "invest", 0.5,
  0.5, -0.62,  "do not", 0.5,
  1.05, 0.9,   "prices settle", 0,
  1.05, -0.9,  "prices settle", 0,
  2.06, 1.35,  "$0.99, $1.22", 0,
  2.06, 0.45,  "$1.19, $1.20", 0,
  2.06, -0.45, "$1.56, $0.94", 0,
  2.06, -1.35, "$1.33, $0.96", 0
)
node <- tibble::tibble(x = c(0, 1, 1), y = c(0, 0.9, -0.9))

ggplot() +
  geom_segment(data = seg, aes(x, y, xend = xend, yend = yend),
               color = itwd_muted, linewidth = 0.7) +
  geom_point(data = node, aes(x, y), size = 3, color = itwd_ink) +
  geom_text(data = lab, aes(x, y, label = t, hjust = h),
            size = 3.5, color = itwd_ink) +
  annotate("text", x = 0.5, y = 1.72, label = "stage 1: commit",
           size = 3.4, color = itwd_muted) +
  annotate("text", x = 1.55, y = 1.72, label = "stage 2: compete on price",
           size = 3.4, color = itwd_muted) +
  coord_cartesian(xlim = c(-0.55, 2.95), ylim = c(-1.85, 1.95)) +
  theme_void()
```

Reading a tree is reading it backwards. The payoffs at the far right are equilibria of the pricing game, computed under whatever was committed at the left. Only once those exist can the first decision be made, which is the backward induction [chapter 16](non-price-competition.qmd) arrived at by necessity rather than by name.

Moving first is not automatically better. A first mover commits while knowing least, and a second mover buys information by waiting and gives up the chance to shape what it is walking into. Which is worth more depends entirely on whether the first move changes anything the second firm has to live with. If it does not, going first is just deciding early.

## The Threat You Cannot Carry Out

Here is where most strategic talk fails.

Suppose Hogi Yogi, tired of losing, announces that it will match any price Smart Cookie sets and undercut it by a nickel. Alarming, if true. Smart Cookie should check whether it is.

Hogi Yogi's variable cost is fifty cents. At Smart Cookie's price of \$1.53, undercutting by a nickel means charging \$1.48 — but Hogi Yogi's own best response to \$1.53 is \$1.02, which earns it more than \$1.48 would. The announced policy is worse for Hogi Yogi than what Hogi Yogi would do anyway, so Hogi Yogi will not do it.

That is a **non-credible threat**: something a firm says it will do that it would not actually want to do when the moment came. Smart Cookie can ignore it, and should.

::: {.column-margin .def-margin}
**Non-credible threat** — a stated intention that the firm would not want to carry out if tested. It changes nothing, because the other side can check the arithmetic as easily as you can.
:::

The test is mechanical and it is the same test throughout this book. Work out what the threatening firm would earn by following through, and what it would earn by quietly not bothering. If the second number is larger, the threat is noise.

Which raises the question the rest of the chapter answers. If saying does not work, what does?

## Commitment Is Removing Your Own Option

A threat becomes credible when carrying it out stops being a sacrifice. The reliable way to arrange that is to change your own situation so the aggressive move is simply what you would now prefer.

That is what commitment is. Not resolve, not stated intention, but an irreversible change to your own payoffs that makes your future behavior predictable to somebody else.

Hogi Yogi has such a move available. Suppose it can install equipment, a better freezer line or a faster assembly process, that cuts its cost per sandwich by fifteen cents, from fifty to thirty-five. The equipment must be bought before any pricing happens, it costs money, and it cannot be returned.

Notice what it does that an announcement cannot. With lower cost, aggressive pricing is not a sacrifice Hogi Yogi is making to punish anyone; it is Hogi Yogi's own best response. Smart Cookie does not have to believe a promise. It has to look at a freezer.

```{r}
#| label: tbl-commitment-game
#| echo: false
#| tbl-cap: "Equipment that cuts unit cost by fifteen cents, available to either firm. Profit per person per month, before the cost of the equipment itself."

eqc <- function(c1, c2) {
  a1 <- -0.7043; b1 <- 2.158877; d1 <- 5.568712
  a2 <- 4.569339; b2 <- 3.496552; d2 <- 0.554085
  X1 <- (a1 + b1*c1)/(2*b1); Y1 <- d1/(2*b1)
  X2 <- (a2 + b2*c2)/(2*b2); Y2 <- d2/(2*b2)
  den <- 1 - Y1*Y2
  p1 <- (X1 + Y1*X2)/den; p2 <- (X2 + Y2*X1)/den
  c((p1-c1)*(a1 - b1*p1 + d1*p2), (p2-c2)*(a2 - b2*p2 + d2*p1))
}
CUT <- 0.15
cc <- function(s,h) { v <- eqc(0.75 - CUT*s, 0.50 - CUT*h)
                      sprintf("%s,  %s", dollar(v[1], .01), dollar(v[2], .01)) }
tibble::tibble(
  ` ` = c("**Smart Cookie does not invest**", "**Smart Cookie invests**"),
  `Hogi Yogi does not invest` = c(cc(0,0), cc(1,0)),
  `Hogi Yogi invests`         = c(cc(0,1), cc(1,1))
) |>
  gt() |> fmt_markdown(columns = ` `) |>
  tab_options(table.font.size = 13, data_row.padding = px(12), table.width = pct(100)) |>
  cols_align("center") |> cols_align("left", columns = ` `)
```

Compare that top-right cell with everything this part has produced so far.

When Hogi Yogi advertised, it gained and Smart Cookie gained more. When it advertised comparatively, the moves cancelled. Here, Hogi Yogi goes from \$0.96 to \$1.22 and *Smart Cookie falls from \$1.33 to \$0.99*. The gap Hogi Yogi has been chasing for three chapters does not narrow — it reverses. Hogi Yogi is now the more profitable firm.

This is the first move in the whole part that takes something from the rival, and the reason is worth sitting with. Reach created customers and the differentiated firm collected most of them. Comparative advertising moved customers and got matched. Lower cost changes what Hogi Yogi will *charge*, and a rival's price is the one thing Smart Cookie's own profit cannot be insulated from.

## The Escalation Nobody Escapes

The equipment is also available to Smart Cookie, so the honest question is the one the last two chapters trained you to ask.

Both firms want it. Smart Cookie gains whether or not Hogi Yogi invests, and so does Hogi Yogi, so long as the equipment costs less than about twenty cents per person per month. At any price below that, both buy, and the equilibrium is the bottom-right cell.

Look at where that leaves them. Before anyone invested, Smart Cookie earned \$1.33 and Hogi Yogi \$0.96. After both invest at, say, twelve cents a head, Smart Cookie earns \$1.07 and Hogi Yogi \$1.08.

Hogi Yogi is better off than it started. Smart Cookie is substantially worse. And Smart Cookie's decision to invest was correct at every step — declining would have left it worse still.

That is not quite a prisoner's dilemma, because only one firm is trapped by it. It is something more common and less discussed: an escalation that transfers advantage from the differentiated firm to the efficient one, in which the differentiated firm's best available play is to participate in its own erosion.

## When Commitment Backfires

The same irreversibility that makes commitment work is what makes it dangerous, and the failures are predictable enough to name.

**It stops being information.** A commitment made on today's estimate of demand is still binding when the estimate turns out wrong. The freezer line does not care that the campus enrollment fell. This is the sunk-cost trap arriving from the front rather than the rear. The problem is not throwing good money after bad; it is having removed your own ability to stop.

**It invites the response it was meant to prevent.** A visible commitment tells your rival exactly what you are about to do, which is the point, and also gives them time to prepare. The table above is that story. Hogi Yogi's move works beautifully against a passive Smart Cookie and much less well against one that reads the same numbers.

**It costs flexibility, and flexibility is worth most exactly when uncertainty is highest.** This is the deepest tension in the chapter, because it puts commitment in direct conflict with everything [Part 3](profit-reasoning.qmd) said about staging decisions under uncertainty. A commitment converts an option into an obligation. When demand is well understood that is cheap; when it is not, you have paid for predictability with your ability to be wrong safely.

There is no formula that resolves that trade. What there is, unexpectedly, is evidence about when it tips.

## What Your Rival Believes

The last chapter said the way out of a trap is to change what your rival expects. This chapter has been about how. It is worth closing on how much that turns out to matter.

Recall the simulated entry decisions from [chapter 15](competitive-advantage.qmd), where each firm held some representation of its market and its rival. Two findings from that work bear directly on commitment.

The first decomposes what makes a rival dangerous to you. A rival that *perceives* well, reading demand accurately and seeing the rivalry it is in, turns out to help you, because accurate perception makes a competitor predictable and rarely reckless. What hurts you is a rival with **commitment capability**: the ability to move irreversibly and make you live with it.^[@hatchCognitiveExposureHow2026, under revision.]

That is a strong statement about the asymmetry of this chapter's subject. Commitment is the attribute of a competitor that damages you most, which is exactly why it is worth having and exactly why it is worth checking whether the firm across the street has it.

The second finding is the one that resolves the tension left open above. The trade between committing and staying flexible is not a fixed property of your situation. *It shifts with the representation your rival holds.* Against a rival who has not thought about you at all, flexibility is cheap and commitment often unnecessary. Against one who is modelling you carefully, moving irreversibly is how you stop being modelled — because there is nothing left to predict.

So the question is not *should I commit or stay flexible*. It is *what does the other firm see, and what does that make worth doing*. Which is the question this entire part has been an argument for asking.

::: {.threshold}
**Before making an irreversible commitment**

- **Is it actually irreversible?** If you could unwind it in a quarter at modest cost, it is a plan rather than a commitment, and it will not change anyone's behavior. That is not a failure; it just means you should not expect the strategic benefit.
- **Does it change what you would want to do, or only what you say?** Compute your best response after the commitment. If the aggressive move still is not your preference, you have bought an expensive announcement.
- **What is the payoff if the rival matches it?** Not whether they will. Fill in the cell. The most common commitment error is evaluating only the column where the rival stands still.
- **What is it worth if your demand estimate is wrong by a third?** Commitments are priced on forecasts and paid for regardless. If the answer is ruinous, the commitment is a bet on the estimate rather than on the strategy.
- **Would you rather face this rival, or be them?** If a competitor made the identical move against you, would you regard it as serious? If not, they will not either.
:::

::: {.reflect}
**Ask yourself — what could you do that you could not undo?**

List the commitments actually available to you in the next year. Equipment, a lease, a hire, an exclusive agreement, a public price, a stated position.

For each one, answer a single question: after making it, what would you *want* to do differently from what you want now? If the answer is nothing, it is a purchase rather than a commitment, and it will not change how anyone else behaves.

Then take the one that does change your incentives and ask the harder version. Your competitor can see it. Given what they will then want to do, are you better off than before you moved?

Most founders evaluate a commitment against the world as it is. The question that matters is how it looks in the world the commitment creates.
:::

::: {.key-action}
**The move:** Commitment works by changing what you will want to do, not by changing what you say. If a move leaves your own best response unchanged, it will not move your rival either, however expensive it was.
:::

This part began by observing that every parameter in the competition chapters was a decision somebody had made. It ends with the sharpest version of that claim: the most consequential decisions are the ones that cannot be revisited, they are made before revenue exists and before the prices they govern are ever charged, and they are evaluated by working out an equilibrium that has not happened yet. Which is the whole argument of this book, arriving at the place where the stakes are highest.
